### 2020-08-22 thesis/thesis.tex
[thesis] Corrections to chapter one

% Disable certain warnings
% chktex-file 18

@@

In this thesis we discuss issues with excessive generality of algebraic effects,
and propose a solution without coalgebraic means along with presentation of
experimental programming language Freak, which is a language with (co) algebraic
effects and (co) handlers, where implementation is based on Continuation Passing
Style translation.

@@

Algebraic effects, while not being a new concept, in the recent years have
received a lot of attention~\cite{effects-bibliography}, both from the theoretical
and practical side.

They give developer necessary tools for declarative programming and writing
programs in a way to strive from exposing implementation details, by defining
computational effects as an API\@ in our code. In the same time, they preserve
a lot of flexibility and have a strong theoretical background.

In fact, that very flexibility can be troublesome. Allowing developer for too
much freedom may lead to undesired behaviour and incorrect programs, for this
reason, we develop type systems, to detect errors before they occur.

Multi-shot resumptions in effect handlers and possibility to drop resumption
are powerful tools to describe fairly complex logic in a concise way. That being
said, they give rise to issues that would not occur in standard control flow,
especially when composing various effects together. Unwanted interaction with
external resources multiple times or not invoking finalisation code are cases
that may occur while using algebraic effects and handlers. Let's consider the
following example:

@@

In the following chapter we provide a background
about effects, define algebraic effects in categorical setting as well as point
out dual coalgebraic effects and cohandlers, finishing with showing related work
in this area. Chapter~\ref{chapter:potential-solutions} describes possible ways
to approach issue of excessive generality,
Chapter~\ref{chapter:co-effectful-programming} shows Freak language by
examples along with usage guide, and then
Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
operational semantics and CPS translation. In next one,
Chapter~\ref{chapter:implementation}, implementation details are revealed.
We conclude in Chapter~\ref{chapter:conclusion} by stating what are the possible
augmentations, that are intended to be made in the future.

@@

    (co) recursive calls& shrinks the argument & grows the result     \\
    \midrule

@@

\chapter{(Co) Effectful Programming}\label{chapter:co-effectful-programming} % TODO w/o space?
\section{Examples}\label{sec:examples}

@@

    Having in mind the drawbacks mentioned above, alternative translation was
    written, that coincides with the translation from~\cite{handlers-cps-journal}.
    Namely, with the uncurried translation to target calculus with continuations
    represented as explicit stacks. The target calculus was described in
    Section~\ref{sec:implementation-ast}. The continuations are represented as
    \verb!Cont!, with syntactic distinction between pure, effectful and
    coeffectful computations. Pure and (co) effectful continuations occupy
    alternating positions in the stack. Explicit distinction was made to provide
    more control in the source code.

@@

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co) effect that occured in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation, where evaluation result is represented by\\
    \verb!type EvalResMonad a = IO (Either Error a)!. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests. Links language~\cite{handlers-cps-journal} also supports
    shallow handlers~\cite{shallow-handlers}, whereas our langauge implements
    only deep ones.

>>>>

% Disable certain warnings
% chktex-file 18
% chktex-file 36

@@

In this thesis we discuss issues arising from combining algebraic effects
and handlers wiht resource management, and present a solution through coalgebraic
means along with presentation of experimental programming language Freak, which
is a language with (co)algebraic effects and (co)handlers, where implementation
is based on Continuation Passing Style translation.

@@

Algebraic effects, while not a new concept, have received a lot of
attention~\cite{effects-bibliography} in recent years, both from the theoretical
and practical side.

They give the developer necessary tools for declarative programming with
computational effects on a high level of abstraction, by defining
effects as an API\@ in the code. At the same time, algebraic effects preserve
a lot of flexibility and have a strong theoretical foundations.

In fact, that very flexibility can be troublesome. Allowing developer for too
much freedom may lead to undesired behaviour and incorrect programs.

In particular, multi-shot resumptions in effect handlers and the possibility to
drop the resumption allow one to express fairly complex logic in a concise way,
but at the same time, they give rise to issues that would not occur in standard
control flow, especially when composing various effects together. Unwanted
interaction with external resources multiple times or not invoking finalisation
code are cases that may occur while using algebraic effects and handlers. Let's
consider the following example:

@@

In the following chapter we provide a background about effects, define algebraic
effects in the categorical setting as well as point out dual coalgebraic effects
and cohandlers, finishing with showing related work in this area.
Chapter~\ref{chapter:potential-solutions} describes possible ways to approach
the issue of excessive generality, Chapter~\ref{chapter:co-effectful-programming}
shows the Freak language by examples along with a usage guide, and then
Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
operational semantics and CPS translation. In the next one,
Chapter~\ref{chapter:implementation}, implementation details are revealed.
We conclude in Chapter~\ref{chapter:conclusion} by stating what the possible
augmentations are, which are intended to be made in the future.

@@

    (co)recursive calls& shrinks the argument & grows the result     \\
    \midrule

@@

\chapter{(Co)Effectful Programming}\label{chapter:co-effectful-programming}
\section{Examples}\label{sec:examples}

@@

    Having in mind the drawbacks mentioned above, alternative translation was
    written, that coincides with the translation from~\cite{handlers-cps-journal}.
    Namely, with the uncurried translation to target calculus with continuations
    represented as explicit stacks. The target calculus was described in
    Section~\ref{sec:implementation-ast}. The continuations are represented as
    \verb!Cont!, with syntactic distinction between pure, effectful and
    coeffectful computations. Pure and (co)effectful continuations occupy
    alternating positions in the stack. Explicit distinction was made to provide
    more control in the source code.

@@

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co)effect that occured in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation, where evaluation result is represented by\\
    \verb!type EvalResMonad a = IO (Either Error a)!. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests. Links language~\cite{handlers-cps-journal} also supports
    shallow handlers~\cite{shallow-handlers}, whereas our langauge implements
    only deep ones.
